When Subgraph Isomorphism is Really Hard, and Why This Matters for Graph Databases

### 2017-10-31 paper/paper.tex
Stuffing

\author{\name Ciaran McCreesh \thanks{This work was supported by the Engineering and Physical Sciences
           Research Council [grant numbers EP/K503058/1, EP/M508056/1, and EP/P026842/1]} \email
           ciaran.mccreesh@glasgow.ac.uk \\
       \addr University of Glasgow, Glasgow, Scotland \\
       \name Patrick Prosser \samethanks[2] \email patrick.prosser@glasgow.ac.uk \\
       \addr University of Glasgow, Glasgow, Scotland \\
       \name Christine Solnon \thanks{This work was supported
       by the ANR project SoLStiCe (ANR-13-BS02-0002-01)} \email christine.solnon@insa-lyon.fr \\
       \addr INSA-Lyon, LIRIS, UMR5205, F-69621, France \\
       \name James Trimble \samethanks[2] \email j.trimble.1@research.gla.ac.uk \\
       \addr University of Glasgow, Glasgow, Scotland}
\maketitle

\begin{abstract}
    The subgraph isomorphism problem involves finding a copy of a pattern graph inside a larger
    target graph. The non-induced version allows extra edges in the target, whilst the induced
    version does not.  Although both subgraph isomorphism variants are \NP-complete, modern
    algorithms can operate comfortably on most real-world graphs with thousands of vertices.
    However, they cannot handle arbitrary instances of this size. We show how to generate ``really
    hard'' random instances for subgraph isomorphism problems, which are computationally challenging
    with only a couple of hundred vertices in the target, and only twenty to thirty pattern
    vertices. For the non-induced version of the problem, these instances lie on a satisfiable /
    unsatisfiable phase transition, whose location we can predict; for the induced variant, much
    richer behaviour is observed, and constrainedness gives a better measure of difficulty than does
    proximity to a phase transition. In both cases, we can use this to understand and improve
    variable- and value-ordering heuristics.

    When labels are introduced, as commonly seen in real-world applications and in graph databases,
    additional rich behaviour arises.  We show that when even the simplest constraint programming
    algorithms are used, all commonly used benchmark suites for evaluating graph database systems
    become trivial, and that existing notions of difficulty are based almost entirely upon
    unnecessarily poor performance from overly simplistic subgraph isomorphism algorithms. With this
    in mind, we demonstrate that the widely researched ``filter~/ verify'' indexing technique used in
    graph databases is a flawed design based upon a repeated misunderstanding of the empirical
    hardness of \NP-complete problems. We explain why filter~/ verify cannot be beneficial, even
    with harder instances than those currently used, when paired with any reasonable modern subgraph
    isomorphism algorithm.
\end{abstract}

@@

The \emph{non-induced subgraph isomorphism problem} is to find an injective mapping from the
vertices of a given pattern graph to the vertices of a given target graph which preserves
adjacency---in essence, we are ``finding a copy of'' the pattern inside the target. The
\emph{induced} variant of the problem additionally requires that the mapping preserve non-adjacency,
so there are no ``extra edges'' in the copy of the pattern that we find. We illustrate both variants
in \cref{figure:sip}.  Although these problems are \NP-complete, modern practical subgraph
isomorphism algorithms can handle problem instances with many hundreds of vertices in the pattern
graph, and up to ten thousand vertices in the target graph
\citep{DBLP:journals/pami/CordellaFSV04,DBLP:journals/ai/Solnon10,DBLP:conf/cp/AudemardLMGP14,DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16},
leading to successful applications in areas such as computer vision
\citep{DBLP:journals/cviu/DamiandSHJS11,DBLP:journals/pr/SolnonDHJ15}, biochemistry
\citep{o:10.1371/journal.pone.0076911,DBLP:conf/gbrpr/CarlettiFV15}, and pattern recognition
\citep{DBLP:journals/ijprai/ConteFSV04}.  In particular, a labelled version of subgraph isomorphism
is one of the key components in supporting complex queries in graph databases, which we discuss in
detail in \cref{section:filterverify}. Typically, these systems store a fixed set of target graphs,
and for a sequence of pattern queries, they must return every target graph which contains that
pattern.  One common approach to this problem is to combine a subgraph isomorphism algorithm with an
indexing system in a so-called ``filter~/ verify'' model, where invariants are used to attempt to
pre-exclude unsatisfiable instances to avoid the cost of a subgraph isomorphism call. In this
context, the terms \emph{matching} and \emph{verification} are often used for the subgraph
isomorphism step (or a slightly broader problem, for example permitting wildcards).

@@

However, subgraph isomorphism algorithms cannot handle \emph{arbitrary} instances involving hundreds
or thousands of vertices. Experimental evaluations of subgraph isomorphism algorithms are usually
performed using randomly generated graph pairs, sometimes together with a mix of real-world graphs,
and graphs that encode biochemistry and computer vision problems. Using random instances to evaluate
algorithm behaviour can be beneficial, because it provides a way of generating many instances
cheaply, and reduces the risk of over-fitting when tuning design parameters. It also allows us to
control graph parameters such as order, density, or degree distribution, and to study scaling
properties of algorithms with respect to these parameters.

The random instances used in previous evaluations typically come from common datasets
\citep{DBLP:journals/prl/SantoFSV03,DBLP:journals/constraints/ZampelliDS10}, which were generated by
taking a random subgraph of a random (Erd\H{o}s-R\'enyi, scale-free, bounded degree, or mesh) graph
and permuting the vertices. Such instances are guaranteed to be satisfiable---\citet{o:AntonO09}
exploited this property to create large sets of random satisfiable boolean satisfiability instances.
This is the most common approach to generating random subgraph isomorphism instances, meaning
existing benchmark suites contain relatively few non-trivial unsatisfiable instances (although a few
of the patterns in the instances by \citeauthor{DBLP:journals/constraints/ZampelliDS10} have had
extra edges added, making them unsatisfiable). Also, the satisfiable instances tend to be
computationally fairly easy, with most of the difficulty being in dealing with the size of the
model. This has lead to bias in algorithm design, to the extent that some proposed techniques, such
as those of \citet{DBLP:conf/sls/BattitiM07}, will \emph{only} work on satisfiable instances.

@@

In \cref{section:filterverify} we look at some of the datasets commonly used to test graph database
systems, which the literature suggests are hard to solve. Simple experiments show that the entire
perceived difficulty of each of these datasets comes down to the widespread use of especially poor
subgraph isomorphism algorithms. Our fourth contribution is to show that the use of forward-checking
and the right variable ordering heuristic, even without the more sophisticated recent advances in
subgraph algorithms, makes the entire graph database filter~/ verify paradigm unnecessary.  Finally,
we explain why filter~/ verify (and other recently proposed techniques) cannot be beneficial even on
more challenging instances except if an extremely poor choice of subgraph isomorphism algorithm is
made.

@@

Part of this paper appeared at IJCAI \citep{DBLP:conf/ijcai/McCreeshPT16};
\cref{section:introduction} has been expanded and
\cref{section:regular,section:labelled,section:filterverify} are new material.

@@

Throughout, our graphs are undirected, and do not have any loops. In some cases vertices have
labels, which are treated as integers; we write $\ell(v)$ for the label of vertex $v$. The
\emph{order} of a graph is the cardinality of its vertex set. We write $\operatorname{V}(G)$ for the
vertex set of a graph $G$. The \emph{complement} of a graph $G$, denoted $\shortoverline{G}$, is the
graph with the same vertex set as $G$, and with an edge between distinct vertices $v$ and $w$ if and
only if $v$ and $w$ are not adjacent in $G$. We write $G(n, p)$ for an Erd\H{o}s-R\'enyi random
graph with $n$ vertices, and an edge between each distinct pair of vertices with independent
probability $p$.

A \emph{non-induced subgraph isomorphism} from a graph $P$ (called the \emph{pattern}) to a graph
$T$ (called the \emph{target}) is an injective mapping from $\operatorname{V}(P)$ to
$\operatorname{V}(T)$ which preserves adjacency---that is, for every adjacent $v$ and $w$ in
$\operatorname{V}(P)$, the vertices $i(v)$ and $i(w)$ are adjacent in $T$. Where labels are present,
the mapping must also preserve labels, so $\ell(v) = \ell(i(v))$ for any $v$. An \emph{induced
subgraph isomorphism} additionally preserves non-adjacency---that is, if $v$ and $w$ are not
adjacent in $P$, then $i(v)$ and $i(w)$ must not be adjacent in $T$. We use the notation $i : P
\rightarrowtail T$ for a non-induced isomorphism, and $i : P \hookrightarrow T$ for an induced
isomorphism. Observe that an induced isomorphism $i : P \hookrightarrow T$ is a non-induced
isomorphism $i : P \rightarrowtail T$ which is also a non-induced isomorphism $\shortoverline{i} :
\shortoverline{P} \rightarrowtail \shortoverline{T}$.

@@

The experiments in this paper are performed on systems with Intel Xeon E5-4650 v2 CPUs, running
Scientific Linux release 6.7. We will be working with four subgraph isomorphism solvers: the Glasgow
solver \citet{DBLP:conf/cp/McCreeshP15}; LAD \citep{DBLP:journals/ai/Solnon10}; VF2
\citep{DBLP:journals/pami/CordellaFSV04}; and VF3 \citep{DBLP:conf/gbrpr/CarlettiFSV17}. Each was
compiled using GCC 4.9.

The Glasgow and LAD solvers store domains, and use backtracking search to build up an assignment of
target vertices (values) to pattern vertices (variables), but differ in terms of inference and
ordering heuristics.  The approach used by VF2 and VF3 is similar, although the domains of variables
are not stored (in the style of conventional backtracking, rather than forward-checking), and so
domain wipeouts are not detected until an assignment is made. Both the Glasgow and LAD solvers use
domain sizes to guide search, whilst VF2 and VF3 use adjacency to previously-assigned vertices.

@@

    \caption{With a fixed pattern graph order of 20, a target graph order of 150, a target edge
        probability of 0.40, and varying pattern edge probability, we observe a phase transition and
        complexity peak with the Glasgow solver in the non-induced variant. Each point represents
        one instance. The lines show mean search effort and mean proportion satisfiable.}
    \label{figure:phase-transition}
\end{figure}

@@

The line plots mean search effort using the Glasgow solver: for sparse patterns, the problem is
trivial, for dense patterns proving unsatisfiability is not particularly difficult, and we see a
complexity peak around the point where half the instances are satisfiable.  We also plot the search
cost of individual instances, as points. The behaviour we observe looks remarkably similar to the
clique decision problem and to random 3SAT problems---compare, for example,
Figure~1 of
\citet{DBLP:journals/cacm/Leyton-BrownHHX14}. In particular, satisfiable instances tend to be easier, but show greater
variation than unsatisfiable instances, and there are exceptionally hard satisfiable instances
\citep{DBLP:conf/cp/SmithG97}. (The Glasgow solver supports
parallel search with a work-stealing strategy explicitly designed to reduce or eliminate these. We
are not using parallel search to avoid dealing with the complexity of search tree
measurements under parallelism.)

@@

\subsection{Variable and Value Ordering Heuristics}

Various general principles have been considered when designing variable and value ordering
heuristics for backtracking search algorithms---one of these is to try to maximise the expected
number of solutions inside any subproblem considered during search \citep{DBLP:conf/cp/GentMPSW96}.
This is usually done by cheaper surrogates, rather than direct calculation. When branching, both LAD
and Glasgow pick a variable with fewest remaining values in its domain: doing this will generally
reduce the first part of the $\langle Sol \rangle$ equation by as little as possible. When two or
more domains are of equal size, LAD simply breaks ties lexicographically, whereas Glasgow will pick
a variable corresponding to a pattern vertex of highest degree (?? check carefully!). This strategy
was determined empirically, but could have been derived from the $\langle Sol \rangle$ formula:
picking a pattern vertex of high degree will make the remaining pattern subgraph sparser, which will
decrease the exponent in the second half of the formula, maximising the overall value. LAD does not
apply a value ordering heuristic, but Glasgow does: it prefers target vertices of highest degree (??
check carefully!).  Again, this was determined empirically, but it has the effect of increasing
$\langle Sol \rangle$ by leaving as many vertices as possible available for future use. The VF2
heuristics, in contrast, are based around preserving connectivity, which gives very little
discrimination except on the sparsest of inputs.

\subsection{Instances for Benchmarking}

?? Instances for benchmarking

?? Something about distance-based filtering being no use on these instances

?? Low degree spread

@@

In the first four rows of \cref{figure:induced} we repeat our experiments, now finding induced
isomorphisms. With a pattern of order 10, we get two independent phase transitions: the bottom right
halves of the plots resemble the non-induced results, and the top left halves are close to a mirror
image. The central satisfiable region, which is away from either phase transition, is
computationally easy, but instances near the phase transition are hard.

For larger patterns of order 20 and 30, we have a large unsatisfiable region in the middle. Despite
not being near either phase transition, instances in the centre remain computationally challenging.
We also plot patterns of orders 14, 15 and 16, to show the transition between the two behaviours.

We might expect these complexity plots to be symmetric along the diagonal, since for the induced
problem, if we replace both inputs with their complements, the solutions remain the same.  For the
Glasgow solver, this is the case. This should be expected, because this complement property is
precisely how the Glasgow solver handles the induced variant (although the heuristics may differ
between the two, which we discuss below). For LAD, some of the very dense patterns are slightly
harder than their diagonal opposites (LAD reasons about degrees, but not about complement-degrees);
it is interesting to note that for larger target graphs, VF2 finds \emph{all} dense pattern graphs
difficult. Finally, VF3's performance generally improves upon VF2's on instances that VF2 finds easy
(except in the cases where VF3 crashes), but worsens on instances that VF2 finds hard.

@@

We might guess that degree-based heuristics would just not work for the induced problem: for any
claim about the degree, the opposite will hold for the complement constraints. However, empirically,
this is not the case: on the final row of \cref{figure:induced}, we show whether it is better to use
the original pattern and target as the input to the Glasgow algorithm, or to take the complements.
(The only steps performed by the Glagsow algorithm which differ under taking the complements are
the degree-based heuristics.  LAD and VF2 are not symmetric in this way: LAD performs a filtering
step using degree information, but does not consider the complement degree, and VF2 uses
connectivity in the pattern graph.)

@@

In practice, this is unlikely to be a problem: most real-world instances are extremely sparse and
are usually easy. In this situation, these experiments justify reusing the
non-induced heuristics on induced problems.

@@

Our pseudo-boolean (PB) encoding is as follows. For each pattern vertex $v$ and each target vertex
$w$, we have a binary variable which takes the value 1 if and only if $v$ is mapped to $w$.
Constraints are added to ensure that each pattern vertex maps to exactly one target vertex, that
each target vertex is mapped to by at most one pattern vertex, that adjacent vertices are mapped to
adjacent vertices, and that non-adjacent vertices are mapped to non-adjacent vertices. We use the
Clasp solver \citep{DBLP:journals/aicom/GebserKKOSS11} version 3.1.3 to solve the pseudo-boolean instances.  The
instances that are hard for the Glasgow solver remain hard for the PB solver, including instances
inside the central region, and the easy satisfiable instances remain easy. Similar results are seen
with the Glucose SAT solver \citep{o:glucose} using a direct encoding of the cardinality constraints.
We also show an integer program encoding: the Gurobi solver is only able to solve some of the
trivial satisfiable instances, and was almost never able to prove unsatisfiability within the time
limit.

@@

Constrainedness, denoted $\kappa$, is an alternative measure of difficulty designed to refine the
phase transition concept, and to generalise hardness parameters across different combinatorial
problems \citep{DBLP:conf/aaai/GentMPW96}. A problem with $\kappa < 1$ is said to be underconstrained, and is
likely to be satisfiable; a problem with $\kappa > 1$ is overconstrained, and is likely to be
unsatisfiable. Empirically, problems with $\kappa$ close to 1 are hard, and problems where $\kappa$
is very small or very large are usually easy. By handling injectivity as a restriction on the size
of the state space rather than as a constraint, we derive
\begin{equation}\label{equation:constrainedness}\kappa = 1 - \frac{\log \left(
t^{\underline{p}} \cdot {d_t}^{d_p \cdot \binom{p}{2}} \cdot {(1 - d_{t})}^{(1 - d_{p}) \cdot
\binom{p}{2}} \right)}{\log t^{\underline{p}}} \end{equation} for induced isomorphisms, which we plot on the
fifth row of \cref{figure:induced}. We see that constrainedness predicts that the central region
will still be relatively difficult for larger pattern graphs: although the problem is
overconstrained, it is less overconstrained than in the regions the Glasgow and LAD solvers found
easy.  Thus it seems that rather than just being a unification of existing generalised heuristic
techniques, constrainedness also gives a better predictor of difficulty than proximity to a phase
transition---our method generates instances where constrainedness and ``close to a phase
transition'' give very different predictions, and constrainedness gives the better prediction.

@@

What about richer graph structures? We now look at what happens with the non-induced problem if we
use $k$-regular graphs rather than the Erd\H{o}s-R\'enyi model, and then in the following section we
see the effects of introducing vertex labels. Both models still exhibit phase transition behaviour,
although with sufficiently many labels the ``satisfiable'' region now contains a mix of satisfiable
and unsatisfiable instances.

@@

?? Talk about degree heuristics

@@

Suppose our labels are drawn randomly from a set $L = \{ 1, \ldots, k \}$, where $k$ is reasonably
small compared to $p$. Recall that $\ell(v)$ is the label of vertex $v$. By defining \[
    \operatorname{V}(P)\vert_n = \{ v \in \operatorname{V}(P) : \ell(v) = n \} \] to be the set of
vertices with label $k$, we can partition the pattern vertices by label into disjoint sets
$\{\operatorname{V}(P)\vert_1, \ldots, \operatorname{V}(P)\vert_k \}\textnormal{\,,}$ each of
which is expected to contain $\nicefrac{p}{k}$ vertices. Similarly, we may partition the target
vertices into disjoint sets $\{ \operatorname{V}(T)\vert_1, \ldots, \operatorname{V}(T)\vert_k \}$.

@@

            \node [anchor=center, inner sep=2pt, rotate=90] {\scriptsize VF2 Failures}; &
            \node [anchor=center] {\includegraphics*{plots/graph-vf2-non-induced-abortions-20-150.pdf}}; &
            \node [anchor=center] {\includegraphics*{plots/graph-vf2-non-induced-abortions-20-l2-150.pdf}}; &

>>>>

\author{\name Ciaran McCreesh \thanks{This work was supported by the Engineering and Physical Sciences
           Research Council [grant numbers EP/K503058/1, EP/M508056/1, and EP/P026842/1]} \email
           ciaran.mccreesh@glasgow.ac.uk \\
       \addr University of Glasgow, Glasgow, Scotland \\
       \name Patrick Prosser \samethanks[1] \email patrick.prosser@glasgow.ac.uk \\
       \addr University of Glasgow, Glasgow, Scotland \\
       \name Christine Solnon \thanks{This work was supported
       by the ANR project SoLStiCe (ANR-13-BS02-0002-01)} \email christine.solnon@insa-lyon.fr \\
       \addr INSA-Lyon, LIRIS, UMR5205, F-69621, France \\
       \name James Trimble \samethanks[1] \email j.trimble.1@research.gla.ac.uk \\
       \addr University of Glasgow, Glasgow, Scotland}
\maketitle

\begin{abstract}
    The subgraph isomorphism problem involves finding a copy of a pattern graph inside a larger
    target graph. The non-induced version allows extra edges in the target, whilst the induced
    version does not.  Although both variants are \NP-complete, algorithms inspired by constraint programming
    can operate comfortably on most real-world graphs with thousands of vertices.  However, they
    cannot handle arbitrary instances of this size. We show how to generate ``really hard'' random
    instances for subgraph isomorphism problems, which are computationally challenging with a couple
    of hundred vertices in the target, and only twenty pattern vertices. For the non-induced version
    of the problem, these instances lie on a satisfiable / unsatisfiable phase transition, whose
    location we can predict; for the induced variant, much richer behaviour is observed, and
    constrainedness gives a better measure of difficulty than does proximity to a phase transition.
    These results have practical consequences: we explain why the widely researched ``filter~/
    verify'' indexing technique used in graph databases is a flawed design based upon a repeated
    misunderstanding of the empirical hardness of \NP-complete problems, that cannot be beneficial
    when paired with any reasonable subgraph isomorphism algorithm.
\end{abstract}

@@

The \emph{non-induced subgraph isomorphism problem} is to find an injective mapping from the
vertices of a given pattern graph to the vertices of a given target graph which preserves
adjacency---in essence, we are ``finding a copy of'' the pattern inside the target. The
\emph{induced} variant of the problem additionally requires that the mapping preserve non-adjacency,
so there are no ``extra edges'' in the copy of the pattern that we find. We illustrate both variants
in \cref{figure:sip}.  Although these problems are \NP-complete, modern subgraph
isomorphism algorithms based upon constraint programming techniques can handle problem instances
with many hundreds of vertices in the pattern graph, and up to ten thousand vertices in the target
graph
\citep{DBLP:journals/ai/Solnon10,DBLP:conf/cp/AudemardLMGP14,DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16},
and subgraph isomorphism is used successfully in application areas including computer vision
\citep{DBLP:journals/cviu/DamiandSHJS11,DBLP:journals/pr/SolnonDHJ15}, biochemistry
\citep{o:10.1371/journal.pone.0076911,DBLP:conf/gbrpr/CarlettiFV15}, and pattern recognition
\citep{DBLP:journals/ijprai/ConteFSV04}.  A labelled version of subgraph isomorphism
is also one of the key components in supporting complex queries in graph databases, which we discuss in
detail in \cref{section:filterverify}.

@@

However, subgraph isomorphism algorithms cannot handle \emph{arbitrary} instances involving hundreds
or thousands of vertices. Experimental evaluations of subgraph isomorphism algorithms are usually
performed using randomly generated graph pairs, sometimes together with a mix of real-world graphs,
and graphs that encode biochemistry and computer vision problems. Using random instances to evaluate
algorithm behaviour can be beneficial, because it provides a way of generating many instances
cheaply, and reduces the risk of over-fitting when tuning design parameters. It also allows
experimenters to control graph parameters such as order, density, or degree distribution, and to
study scaling properties of algorithms with respect to these parameters.  The random instances used
in previous evaluations come from common datasets
\citep{DBLP:journals/prl/SantoFSV03,DBLP:journals/constraints/ZampelliDS10}, which were generated by
taking a random subgraph of a random (Erd\H{o}s-R\'enyi, scale-free, bounded degree, or mesh) graph
and permuting the vertices. Such instances are guaranteed to be satisfiable---\citet{o:AntonO09}
exploited this property to create large sets of random satisfiable boolean satisfiability instances.
This is the most common approach to generating random subgraph isomorphism instances, meaning
existing benchmark suites contain relatively few non-trivial unsatisfiable instances (although a few
of the patterns in the instances by \citeauthor{DBLP:journals/constraints/ZampelliDS10} have had
extra edges added, making them unsatisfiable). Also, the satisfiable instances tend to be
computationally fairly easy, with most of the difficulty being in dealing with the size of the
model. This has lead to bias in algorithm design, to the extent that some proposed techniques, such
as those of \citet{DBLP:conf/sls/BattitiM07}, will \emph{only} work on satisfiable instances.

@@

In \cref{section:filterverify} we look at some of the datasets commonly used to test graph database
systems, which the literature suggests are hard to solve. Simple experiments show that the entire
perceived difficulty of each of these datasets comes down to the widespread use of especially poor
subgraph isomorphism algorithms. Our fourth contribution is to show that the use of forward-checking
and the right variable-ordering heuristic, even without the more sophisticated recent advances in
subgraph algorithms, makes the entire graph database filter~/ verify paradigm unnecessary.  Finally,
we explain why filter~/ verify (and other recently proposed techniques, which we discuss in
\cref{section:angryranting}) cannot be beneficial even on more challenging instances except if an
extremely poor choice of subgraph isomorphism algorithm is made.

@@

Part of this paper were presented at IJCAI \citep{DBLP:conf/ijcai/McCreeshPT16}; the first three
sections have been expanded, whilst
\cref{section:regular,section:labelled,section:filterverify,section:angryranting} are new material.

@@

Throughout, our graphs are undirected, and do not have any loops. In some cases vertices have
labels, which are treated as integers; we write $\ell(v)$ for the label of vertex $v$. The
\emph{order} of a graph is the cardinality of its vertex set. We write $\operatorname{V}(G)$ for the
vertex set of a graph $G$. The \emph{complement} of a graph $G$, denoted $\shortoverline{G}$, is the
graph with the same vertex set as $G$, and with an edge between distinct vertices $v$ and $w$ if and
only if $v$ and $w$ are not adjacent in $G$. We write $G(n, p)$ for an Erd\H{o}s-R\'enyi random
graph with $n$ vertices, and an edge between each distinct unordered pair of vertices with
independent probability $p$.

A \emph{non-induced subgraph isomorphism} from a graph $P$ (called the \emph{pattern}) to a graph
$T$ (called the \emph{target}) is an injective mapping $i$ from $\operatorname{V}(P)$ to
$\operatorname{V}(T)$ which preserves adjacency---that is, for every adjacent $v$ and $w$ in
$\operatorname{V}(P)$, the vertices $i(v)$ and $i(w)$ are adjacent in $T$. Where labels are present,
the mapping must also preserve labels, so $\ell(v) = \ell(i(v))$ for any $v$. An \emph{induced
subgraph isomorphism} additionally preserves non-adjacency---that is, if $v$ and $w$ are not
adjacent in $P$, then $i(v)$ and $i(w)$ must not be adjacent in $T$. We use the notation $i : P
\rightarrowtail T$ for a non-induced isomorphism, and $i : P \hookrightarrow T$ for an induced
isomorphism. Observe that an induced isomorphism $i : P \hookrightarrow T$ is a non-induced
isomorphism $i : P \rightarrowtail T$ which is also a non-induced isomorphism $\shortoverline{i} :
\shortoverline{P} \rightarrowtail \shortoverline{T}$.

@@

The experiments in this paper are performed on systems with Intel Xeon E5-4650 v2 CPUs, running
Scientific Linux release 6.7. We will be working with four subgraph isomorphism solvers: the Glasgow
solver \citep{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16}; LAD
\citep{DBLP:journals/ai/Solnon10}; VF2 \citep{DBLP:journals/pami/CordellaFSV04}; and VF3
\citep{DBLP:conf/gbrpr/CarlettiFSV17}. Each was compiled using GCC 4.9.

The Glasgow and LAD solvers are constraint programming style algorithms (although they use dedicated
data structures, and do not use a constraint programming toolkit). For each vertex in the pattern
graph, they create a \emph{variable} whose \emph{domain} contains one \emph{value} for each vertex
in the target graph. The algorithms perform backtracking search, with the goal of assigning to each
variable a value from its domain so as to form a subgraph isomorphism. The algorithms use
\emph{inference} to process \emph{constraints} which eliminate values from domains, ensuring that
the injectivity (an \emph{all-different} constraint) and adjacency rules are respected. If at any
point during search a \emph{domain wipeout} occurs (that is, if every value is removed from any
domain), the algorithms will backtrack immediately. The search is guided by \emph{variable-ordering
heuristics}, which select which variable (i.e.\ pattern vertex) to branch on next, and
\emph{value-ordering} heuristics which select which value (i.e.\ target vertex) to try as part of a
partial mapping. The two algorithms differ in terms of additional inference performed, their
choices of underlying data structures and algorithms, and in the details of their variable- and
value-ordering heuristics.

The approach used by VF2 and VF3 is a little different: domains of variables are not stored (so the
algorithm operates in the style of conventional backtracking), and wipeouts are not detected until
an assignment is made. Instead, an assignment is built up by repeatedly branching only on vertices
which are adjacent to a previously-assigned vertex.

@@

    \caption{With a fixed pattern graph order of 20, a target graph order of 150, a target edge
        probability of 0.40, and varying pattern edge probability, we observe a phase transition and
        complexity peak with the Glasgow solver in the non-induced variant. Each point represents
        one instance, and the line shows mean search effort.}
    \label{figure:phase-transition}
\end{figure}

@@

The line plots mean search effort using the Glasgow solver: for sparse patterns, the problem is
trivial, for dense patterns proving unsatisfiability is not particularly difficult, and we see a
complexity peak around the point where half the instances are satisfiable.  We also plot the search
cost of individual instances, as points. The behaviour we observe looks remarkably similar to random
3SAT problems---compare, for example, Figure~1 of \citet{DBLP:journals/cacm/Leyton-BrownHHX14}. In
particular, satisfiable instances tend to be easier, but show greater variation than unsatisfiable
instances, and there are exceptionally hard satisfiable instances \citep{DBLP:conf/cp/SmithG97}.

@@

\subsection{Variable- and Value-Ordering Heuristics}

Various general principles have been considered when designing variable- and value-ordering
heuristics for backtracking search algorithms---one of these is to try to maximise the expected
number of solutions inside any subproblem considered during search \citep{DBLP:conf/cp/GentMPSW96}.
This is usually done by cheaper surrogates, rather than direct calculation. When branching, both LAD
and Glasgow pick a variable with fewest remaining values in its domain: doing this will generally
reduce the first part of the $\langle Sol \rangle$ equation by as little as possible. When two or
more domains are of equal size, LAD simply breaks ties lexicographically, whereas Glasgow will pick
a variable corresponding to a pattern vertex of highest degree. This strategy was determined
empirically, but could have been derived from the $\langle Sol \rangle$ formula: picking a pattern
vertex of high degree will make the remaining pattern subgraph sparser, which will decrease the
exponent in the second half of the formula, maximising the overall value. LAD does not apply a
value-ordering heuristic, but Glasgow does: it prefers target vertices of highest degree.  Again, this was
determined empirically, but it has the effect of increasing $\langle Sol \rangle$ by leaving as many
vertices as possible available for future use. The search strategy used by VF2, in contrast, is
based around preserving connectivity, which gives very little discrimination except on the sparsest
of inputs.

@@

So far we have looked at the non-induced problem, where extra edges in the target graph are
permitted.  In the first five rows of \cref{figure:induced} we repeat our experiments, now finding
induced isomorphisms. With a pattern of order 10, we get two independent phase transitions: the
bottom right halves of the plots resemble the non-induced results, and the top left halves are close
to a mirror image. The central satisfiable region, which is away from either phase transition, is
computationally easy, but instances near the phase transition are hard.

For larger patterns of order 20 and 30, we have a large unsatisfiable region in the middle. Despite
not being near either phase transition, instances in the centre remain computationally challenging.
We also plot patterns of orders 14, 15 and 16, to show the change between the two behaviours.

We might expect these complexity plots to be symmetric along the diagonal, since for the induced
problem, if we replace both inputs with their complements, the solutions remain the same.  For the
Glasgow solver, this is almost the case. This should be expected, because this complement property is
precisely how the Glasgow solver handles the induced variant (although the heuristics may differ
between the two, which we discuss below). For LAD, some of the very dense patterns are slightly
harder than their diagonal opposites (LAD reasons about degrees, but not about complement-degrees);
it is interesting to note that for larger target graphs, VF2 finds \emph{all} dense pattern graphs
difficult. Finally, VF3's performance generally improves upon VF2's on instances that VF2 finds easy
(except in the cases where VF3 crashes), but worsens on instances that VF2 finds hard.

@@

We might guess that degree-based heuristics would just not work for the induced problem: for any
claim about the degree, the opposite will hold for the complement constraints. Should we therefore
resort to just using ``smallest domain first'', abandoning degree-based tiebreaking and the
value-ordering heuristic? Empirically, this is not the case: on the final row of
\cref{figure:induced}, we show whether it is better to use the original pattern and target as the
input to the Glasgow algorithm, or to take the complements.  (The only steps performed by the
Glagsow algorithm which differ under taking the complements are the degree-based heuristics.  LAD,
VF2, and VF3 are not symmetric in this way: LAD performs a filtering step using degree information,
but does not consider the complement degree, and VF2 and VF3 use connectivity.)

@@

In practice, this is unlikely to be a problem: most real-world instances are relatively sparse. In
this situation, these experiments justify reusing the non-induced heuristics on induced problems.

@@

The pseudo-boolean (PB) encoding is as follows. For each pattern vertex $v$ and each target vertex
$w$, we have a binary variable which takes the value 1 if and only if $v$ is mapped to $w$.
Constraints are added to ensure that each pattern vertex maps to exactly one target vertex, that
each target vertex is mapped to by at most one pattern vertex, that adjacent vertices are mapped to
adjacent vertices, and that non-adjacent vertices are mapped to non-adjacent vertices. We use the
Clasp solver \citep{DBLP:journals/aicom/GebserKKOSS11} version 3.1.3 to solve the pseudo-boolean instances.  The
instances that are hard for the Glasgow solver remain hard for the PB solver, including instances
inside the central region, and the easy satisfiable instances remain easy. Similar results are seen
with the Glucose SAT solver \citep{o:glucose} using a direct encoding.  We also show an integer
program encoding: the Gurobi solver is only able to solve some of the trivial satisfiable instances,
and was almost never able to prove unsatisfiability within the time limit.

@@

Constrainedness, denoted $\kappa$, is an alternative measure of difficulty designed to refine the
phase transition concept, and to generalise hardness parameters across different combinatorial
problems \citep{DBLP:conf/aaai/GentMPW96}. A problem with $\kappa < 1$ is said to be underconstrained, and is
likely to be satisfiable; a problem with $\kappa > 1$ is overconstrained, and is likely to be
unsatisfiable. Empirically, problems with $\kappa$ close to 1 are hard, and problems where $\kappa$
is very small or very large are usually easy. By handling injectivity as a restriction on the size
of the state space rather than as a constraint, we derive
\begin{equation}\label{equation:constrainedness}\kappa = 1 - \frac{\log \left(
t^{\underline{p}} \cdot {d_t}^{d_p \cdot \binom{p}{2}} \cdot {(1 - d_{t})}^{(1 - d_{p}) \cdot
\binom{p}{2}} \right)}{\log t^{\underline{p}}} \end{equation} for induced isomorphisms, which we plot on the
sixth row of \cref{figure:induced}. We see that constrainedness predicts that the central region
will still be relatively difficult for larger pattern graphs: although the problem is
overconstrained, it is less overconstrained than in the regions the Glasgow and LAD solvers found
easy.  Thus it seems that rather than just being a unification of existing generalised heuristic
techniques, constrainedness also gives a better predictor of difficulty than proximity to a phase
transition---our method generates instances where constrainedness and ``close to a phase
transition'' give very different predictions, and constrainedness gives the better prediction.

@@

What about richer graph structures? We briefly look at what happens with the non-induced problem if
we use $k$-regular graphs rather than the Erd\H{o}s-R\'enyi model, and then in the following section
we see the effects of introducing vertex labels. Both models still exhibit phase transition
behaviour, although with sufficiently many labels the ``satisfiable'' region now contains a mix of
satisfiable and unsatisfiable instances.

@@

Regularity means that degree-based heuristics have no information to work with. Additionally, the
degree-based filtering techniques used by Glasgow and LAD have no effect on these graphs, and so at
the top of search, every domain is identical. However, once a guessed assignment has been made,
selecting from small domains first remains an effective strategy to guide the remainder of the
search.

@@

Suppose our labels are drawn randomly from a set $L = \{ 1, \ldots, k \}$, where $k$ is reasonably
small compared to $p$. Recall that $\ell(v)$ is the label of vertex $v$. By defining \[
    \operatorname{V}(P)\vert_n = \{ v \in \operatorname{V}(P) : \ell(v) = n \} \] to be the set of
vertices with label $n$, we can partition the pattern vertices by label into disjoint sets
$\{\operatorname{V}(P)\vert_1, \ldots, \operatorname{V}(P)\vert_k \}\textnormal{\,,}$ each of
which is expected to contain $\nicefrac{p}{k}$ vertices. Similarly, we may partition the target
vertices into disjoint sets $\{ \operatorname{V}(T)\vert_1, \ldots, \operatorname{V}(T)\vert_k \}$.

@@

            % \node [anchor=center, inner sep=2pt, rotate=90] {\scriptsize VF2 Failures}; &
            % \node [anchor=center] {\includegraphics*{plots/graph-vf2-non-induced-abortions-20-150.pdf}}; &
            % \node [anchor=center] {\includegraphics*{plots/graph-vf2-non-induced-abortions-20-l2-150.pdf}}; &
